---
title: "Mean Citedness analysis"
output:
  html_fragment:
    self_contained: false
    fig_width: 8
    fig_height: 5
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.path = "../data/impact_factor_figs/", dev = "png", dpi = 110
)
```

```{r load}
suppressPackageStartupMessages({
  library(jsonlite)
  library(ggplot2)
  library(scales)
  library(patchwork)
  library(mgcv)
  library(pROC)
  library(lme4)
})

FLoRA_studies <- read.csv("../data/flora_with_omc.csv",
                         stringsAsFactors = FALSE, na.strings = c("", "NA"))

# Disciplines lookup from data/disciplines.json — single source of truth
DISCIPLINES <- fromJSON("../data/disciplines.json", simplifyVector = FALSE)
normalize <- function(x) {
  x <- tolower(x)
  x <- gsub("[^a-z0-9 ]", " ", x)
  x <- gsub("\\s+", " ", x)
  trimws(x)
}
lookup <- do.call(rbind, lapply(names(DISCIPLINES), function(disc) {
  data.frame(journal_norm = normalize(unlist(DISCIPLINES[[disc]])),
             discipline   = disc, stringsAsFactors = FALSE)
}))
FLoRA_studies$journal_norm <- normalize(FLoRA_studies$journal_o)
FLoRA_studies$discipline   <- lookup$discipline[match(FLoRA_studies$journal_norm, lookup$journal_norm)]
FLoRA_studies$discipline   <- factor(ifelse(is.na(FLoRA_studies$discipline),
                                            "Uncategorized", FLoRA_studies$discipline))
```

```{r filter}
# Restrict to replications with binary outcomes, drop extreme OMC > 35
FLoRA_studies$impact_factor <- suppressWarnings(as.numeric(FLoRA_studies$impact_factor))
FLoRA_studies <- FLoRA_studies[!is.na(FLoRA_studies$impact_factor) & FLoRA_studies$impact_factor < 35, ]
FLoRA_studies <- FLoRA_studies[tolower(FLoRA_studies$outcome) %in% c("successful", "failed"), ]
FLoRA_studies$outcome_binary <- ifelse(tolower(FLoRA_studies$outcome) == "failed", 0, 1)
FLoRA_studies$omc_log <- log(FLoRA_studies$impact_factor + 1)
FLoRA_studies$outcome_fac <- factor(tolower(FLoRA_studies$outcome),
                                    levels = c("failed", "successful"))
```

```{r empty-bail, results="asis"}
if (nrow(FLoRA_studies) < 30) {
  cat("<p><em>Not enough studies with both a binary outcome and a journal-level OMC available yet. ",
      "This analysis will populate as more journals are matched to OpenAlex sources.</em></p>\n")
  knitr::knit_exit()
}
```

```{r bt-check}
FLoRA_bt <- FLoRA_studies[FLoRA_studies$omc_log > 0, ]
FLoRA_bt$bt_log <- FLoRA_bt$omc_log * log(FLoRA_bt$omc_log)
bt_log_FLoRA <- glm(outcome_binary ~ omc_log + bt_log,
                    family = binomial, data = FLoRA_bt)
bt_FLoRA_sig <- summary(bt_log_FLoRA)$coefficients["bt_log", "Pr(>|z|)"] < 0.05
```

```{r models}
lr_FLoRA <- glm(outcome_binary ~ omc_log, family = binomial, data = FLoRA_studies)
chi_FLoRA   <- lr_FLoRA$null.deviance - lr_FLoRA$deviance
chidf_FLoRA <- lr_FLoRA$df.null - lr_FLoRA$df.residual
chip_FLoRA  <- 1 - pchisq(chi_FLoRA, chidf_FLoRA)

gam_FLoRA <- gam(outcome_binary ~ s(omc_log), family = binomial, data = FLoRA_studies)

null_FLoRA <- glm(outcome_binary ~ 1, family = binomial, data = FLoRA_studies)
r2_lr_FLoRA  <- as.numeric(1 - logLik(lr_FLoRA)  / logLik(null_FLoRA))
r2_gam_FLoRA <- as.numeric(1 - logLik(gam_FLoRA) / logLik(null_FLoRA))
auc_lr_FLoRA <- as.numeric(auc(roc(lr_FLoRA$y,  fitted(lr_FLoRA),  quiet = TRUE)))
auc_gam_FLoRA <- as.numeric(auc(roc(gam_FLoRA$y, predict(gam_FLoRA, type = "response"), quiet = TRUE)))

b0_f <- coef(lr_FLoRA)[1]; b1_f <- coef(lr_FLoRA)[2]
thr_FLoRA <- if (abs(b1_f) > 1e-9) exp(-b0_f / b1_f) - 1 else NA_real_
```

```{r multilevel, results="hide"}
ml_null  <- tryCatch(glmer(outcome_binary ~ 1 + (1 | discipline),
                           family = binomial, data = FLoRA_studies), error = function(e) NULL)
ml_omc   <- tryCatch(glmer(outcome_binary ~ omc_log + (1 | discipline),
                           family = binomial, data = FLoRA_studies), error = function(e) NULL)
icc_f <- beta_f <- pval_f <- NA_real_
if (!is.null(ml_omc)) {
  tau2_f <- as.numeric(VarCorr(ml_omc)$discipline)
  icc_f  <- tau2_f / (tau2_f + (pi^2 / 3))
  beta_f <- fixef(ml_omc)["omc_log"]
  pval_f <- summary(ml_omc)$coefficients["omc_log", "Pr(>|z|)"]
}
```

## Descriptive Overview

The distribution of OMC values for the FLoRA replications with binary
outcomes is shown below (studies with `OMC > 35` are excluded).

```{r hist, fig.cap="Absolute count of replication outcomes by OMC in the FLoRA database. Studies with OMC > 35 are excluded."}
ggplot(FLoRA_studies, aes(x = impact_factor, fill = outcome_fac)) +
  geom_histogram(position = "stack", binwidth = 0.5,
                 alpha = 0.85, color = "white", linewidth = 0.1) +
  scale_fill_manual(values = c("failed" = "#e05d5d", "successful" = "#48bb78"),
                    labels = c("failed" = "Failed", "successful" = "Successful"),
                    name = "Replication Outcome") +
  coord_cartesian(xlim = c(0, 20)) +
  labs(x = "OpenAlex Mean Citedness (OMC)", y = "Study Count") +
  theme_minimal(base_size = 11)
```

## Results

```{r results-table, results="asis"}
cat(sprintf(
"<p><strong>N = %d</strong> replications with both an outcome and an OMC value (%d successful, %d failed). Logistic regression with log-transformed OMC gave a statistically significant overall model fit, $\\chi^2$(%d) = %.2f, %s. The Box–Tidwell test of linearity-in-log-odds was %s after log-transformation.</p>",
nrow(FLoRA_studies),
sum(FLoRA_studies$outcome_binary == 1), sum(FLoRA_studies$outcome_binary == 0),
chidf_FLoRA, chi_FLoRA,
ifelse(chip_FLoRA < 0.001, "p < .001", sprintf("p = %.3f", chip_FLoRA)),
ifelse(bt_FLoRA_sig, "violated (a GAM is preferred for inference)",
                    "satisfied")
))

cat(sprintf(
"<p>Model performance is modest: McFadden's $R^2$ = %.3f (logistic regression) and %.3f (GAM); AUC = %.2f and %.2f respectively. The OMC at which P(success) = 0.50 (logistic model) is %s.</p>",
r2_lr_FLoRA, r2_gam_FLoRA, auc_lr_FLoRA, auc_gam_FLoRA,
ifelse(is.na(thr_FLoRA), "not estimable", sprintf("%.2f", thr_FLoRA))
))
```

```{r curves, fig.cap="Predicted probability of successful replication as a function of OMC. Left: logistic regression with log-transformed OMC. Right: GAM. Shaded bands represent 95% confidence intervals."}
pred_f <- data.frame(
  impact_factor = seq(min(FLoRA_studies$impact_factor, na.rm = TRUE),
                      max(FLoRA_studies$impact_factor, na.rm = TRUE),
                      length.out = 500))
pred_f$omc_log <- log(pred_f$impact_factor + 1)

lp_f <- predict(lr_FLoRA, newdata = pred_f, type = "link", se.fit = TRUE)
pred_f$p_lr  <- plogis(lp_f$fit)
pred_f$lo_lr <- plogis(lp_f$fit - 1.96 * lp_f$se.fit)
pred_f$hi_lr <- plogis(lp_f$fit + 1.96 * lp_f$se.fit)

gp_f <- predict(gam_FLoRA, newdata = pred_f, type = "link", se.fit = TRUE)
pred_f$p_gam  <- plogis(gp_f$fit)
pred_f$lo_gam <- plogis(gp_f$fit - 1.96 * gp_f$se.fit)
pred_f$hi_gam <- plogis(gp_f$fit + 1.96 * gp_f$se.fit)

jitter_f <- geom_jitter(data = FLoRA_studies,
                        aes(x = impact_factor, y = outcome_binary),
                        alpha = 0.15, height = 0.03, width = 0, inherit.aes = FALSE)

p_lr <- ggplot(pred_f, aes(x = impact_factor)) +
  geom_ribbon(aes(ymin = lo_lr, ymax = hi_lr), fill = "#6b4582", alpha = 0.18) +
  geom_line(aes(y = p_lr), color = "#6b4582", linewidth = 1) +
  jitter_f +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray40") +
  { if (!is.na(thr_FLoRA))
      list(geom_vline(xintercept = thr_FLoRA, linetype = "dashed",
                      color = "#e05d5d", linewidth = 0.8),
           annotate("text", x = thr_FLoRA + 0.4, y = 0.43,
                    label = sprintf("OMC = %.2f", thr_FLoRA),
                    color = "#e05d5d", hjust = 0, size = 3))
    else NULL } +
  scale_y_continuous(limits = c(0, 1), labels = percent_format()) +
  labs(title = "Logistic Regression", x = "OMC",
       y = "P(successful replication)", caption = "shaded: 95% CI") +
  theme_minimal(base_size = 11)

p_gam <- ggplot(pred_f, aes(x = impact_factor)) +
  geom_ribbon(aes(ymin = lo_gam, ymax = hi_gam), fill = "#6b4582", alpha = 0.18) +
  geom_line(aes(y = p_gam), color = "#6b4582", linewidth = 1) +
  jitter_f +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray40") +
  scale_y_continuous(limits = c(0, 1), labels = percent_format()) +
  labs(title = "GAM", x = "OMC", y = NULL, caption = "shaded: 95% CI") +
  theme_minimal(base_size = 11)

p_lr | p_gam
```

```{r discipline, results="asis"}
if (!is.na(icc_f)) {
  cat(sprintf(
"<p><strong>Role of scientific discipline.</strong> A multilevel logistic regression with a random intercept for discipline yields ICC = %.3f, meaning %.1f%% of the variance in replication outcomes lies between disciplines. After accounting for that clustering, log-OMC retains a fixed effect of $\\hat\\beta$ = %.2f (%s).</p>",
icc_f, 100 * icc_f, beta_f,
ifelse(pval_f < 0.001, "p < .001", sprintf("p = %.3f", pval_f))))
}
```

## Discussion

Across model specifications, higher OMC is weakly negatively associated
with replication success. The effect is small in magnitude and the
explanatory power is too limited for journal-level citation metrics to
serve as a meaningful heuristic for judging the replicability of
individual studies.
